Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-02-07 analysis/paper_draft3.Rmd
draft with updated Discussion section

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make implicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection process. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences the bias detection (whether the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected such historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2022.

Without historical strike data, there is a higher risk of underpowered or inflated estimates of attorney strike bias, because of the low number of strikes per trial (typically 6, 10, or 15 strikes per party, depending on the jurisdiction) [@nscs-jury2018]. Yet, completely pooling available current and past strike data ignores possible incompatibility between current and historical strike data. Such incompatibility might arise due to missingness in the historical data or because an attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity).

As a result, incompatibility might cause an attorney's bias parameter based on strike data for an attorney in the current trial to differ from such values based on historical strike data for that attorney. This can lead to inflated type I error and reduced power [@Bennett2021]. To increase power, one may use other methods, such as stratifying strike data by trial and applying the Cochran-Mantel-Haenszel (CMH) test [@Gastwirth2014, p. 295-297]. Such approaches, however, often adjust for incompatibility implicitly. For example, applying the CMH test requires first selecting only historical strike data from the subset of past trials that are "similar" in its bias-salient characteristics to the current trial [@Gastwirth2014, p. 289, 297], a step to reduce incompatibility due to variation in those salient characteristics.

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows for transparently adjusting for the assumed degree of incompatibility between current and historical trials. For this approach, we proceed in three steps.  First, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Second, we use an attorney's strike data from past trials, generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Chen2000; @Ibrahim2015], which raises the likelihood of the historical strike data to a fixed power ($\alpha_{0}$) between zero and one, to weight the historical strike data in accord with expected compatibility between the historical trials and the current trial. Zero denotes no weight to the historical data (complete incompatibility) and one denotes equal weight (complete compatibility). Third, we use this posterior distribution as an informative prior for the bias parameter in the current trial.

Here, we demonstrate this approach with simulations in which we vary the degree of incompatibility between current and historical trial data, and test how well we can "detect" attorney strike bias (i.e., whether the credible interval for the bias parameter excludes zero), given different values of the power prior's alpha parameter. Then, we present a software prototype that encodes the same approach with actual historical strike data. With that data, the prototype shows how attorneys and others can use this approach to detect strike bias in real time during jury selection.

To be sure, neither this nor any other statistical approach can automatically detect _Batson_ violations, because _Batson_ law permits any evidence of strike bias, not just the attorney's strikes in past cases. And it requires the trial judge to weigh all such evidence to decide whether strike bias likely exists (the _Batson_ violation). Still, _before_ accounting for other relevant evidence of strike bias, our approach can help attorneys "detect" strike bias, given current and historical strike data, the model, and a fixed value for the power prior's $\alpha$ parameter. After testing for sensitivity of bias detection to different fixed $\alpha$ values [@Carvalho2021, p. 5252; @Ibrahim2015, p. 3734], attorneys can combine the resulting inference with other evidence to decide whether to argue _Batson_; to prove the _Batson_ prima facie case; and to persuade the trial judge to find  a _Batson_ violation (Figure \@ref(fig:batsonworkflowfig)).

@@

Following others [e.g., @Kadane2018a; @Barrett2007; @Gastwirth2005, 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

To evaluate the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. <!--To measure convergence, we used trace plots, the Gelman-Rubin convergence diagnostic and Geweke’s diagnostics [@Gelman1992;@Gelman1998;@Geweke1992].--> These simulations primarily show how much the size of the historical data affects our ability to detect bias and how much the power prior discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

@@

Bias detection will also be sensitive to the choice of the standard deviation of the initial prior ($\sigma$). The results above depend in part on our choice of a conservative initial prior for the bias parameter (Equation \@ref(eq:initialprior)). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, the model we use will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. As in the simulations, a normal distribution was assumed for the distribution of $b$. The length of the Markov chain was 110000 and the first 10000 iterations were dropped as burn-in. No thinning was performed, as the correlation was weak and convergence occured rapidly. Convergence was checked from both traceplots and the Gelman-Rubin statistic [@Gelman1992]. Computation times varies from 0.76 to 1.2 seconds, depending in the amount of data used (quicker with current data only _versus_ current plus historical data).

To show how lawyers might use this app in real cases in real time during selection, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. Using publicly-available court docket sheets for each case, we matched particular lawyers to these strikes and later de-identified them. For details on this dataset, see the Appendix.

@@

To illustrate, suppose the strike tally in Table 1 depicts the pattern of strikes in the present case with defense attorney Aaron Waelchi and prosecutor Lawrance Klocko V (both aliases for actual attorneys in the historical strike data). After the user enters this strike tally, the app initially displays two graphs -- one for the prosecution and the defense. Each graph depicts the prior density plot (colored light grey) and posterior density plot (blue and red for defense and prosecution, respectively) for the bias parameter (Figure \@ref(fig:figapp1)). Here, the 80\% credible interval includes zero, indicating no credible basis to infer bias, given the current strike tally.

```{r figapp1, out.width='75%', fig.cap = "Screenshot of R-Shiny application showing density plots for bias of prosecutor and defense attorney based on data in 'Current strike tally' table. Vertical dotted lines depict 80 pct. credible interval.", fig.align='center'}

load_path <- here::here("figures","batson_app_screenshots", "batson_app_screenshot1.png")
knitr::include_graphics(load_path)

@@

In our illustration, we select the names of the prosecutor and defense attorney from their respective pull-down menus; and leave the weight setting to "Equal". The density plots update accordingly (Figure \@ref(fig:figapp2)).

```{r figapp2, out.width='75%', fig.cap = "Density Plots with Historical Strike Data (Race). Screenshot of R-Shiny application show density plots based on historical and current strike data for prosecutor and defense attorney.", fig.align='center'}

load_path <- here::here("figures","batson_app_screenshots", "batson_app_screenshot2.png")

@@

Now, the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case. In the Appendix, we present additional illustrations wherein we use the app to detect gender bias and where we assign less weight to the historical strike data.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  prove illegal bias alone, but rather to lend support to an inference of bias and a subsequent _Batson_ challenge. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

Moreover, with a tool like our prototype app, attorneys can use the approach presented here to better detect strike bias for purposes of bringing or satisfying proof burden at different stages of _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence, along with any other admissible evidence, for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)).

To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail. The approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike.

Moreover, suppose an attorney of interest exercised strikes in two courts, each with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that can be applied to another court's procedure (e.g., the simultaneous strike example in section \@ref(simul-strikes) ). In other cases, however, one cannot apply the same model to courts with different jury selection practices. For example, the model we used here does not apply to strike data generated by a court's strike procedure in which prospective jurors appear before the attorneys one at a time to be either struck or seated on the jury.

Furthermore, the bias detection rate depends on the degree of incompatibility between the bias parameters for the current trial and for the past (historical) trials ($\lvert b_{curr} - b_{hist} \rvert$). These bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

First, such incompatibility may occur because of missing historical data. Historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an obstacle for any approach to detecting strike bias, including but not limited to the one here, if the missing data is missing not at random.

US law currently affords a way to reduce this missing data problem. The public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public. Similarly, a criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors [@press1984; @presley2010]. If these rights are asserted, any person can often attend jury voir dire and collect relevant strike data based on what they observe in real time.  Thus, though historical strike data is hard to collect retrospectively, such courtroom observers can prospectively build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

Second, even with complete strike data (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity) that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial with a Black defendant may depend in part on whether we use all the historical data or only the subset of past trials with Black defendants.

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select. Given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

Bias detection also increases with the number of historical trials and the number of strikes per trial. Higher values of $\alpha$ lead to higher bias detection rates across compatibility scenarios, especially in scenarios with a lower number of strikes per trial and few historical trials (top left panel of Figure \@ref(fig:figcmptbly)). The choice of $\alpha$ matters less for bias detection when bias is high; the number of strikes per trial is high; and the number of historical trials is high (lower right panel of Figure \@ref(fig:figcmptbly)). Overall, given high compatibility, increasing $\alpha$ will increase bias detection.

In cases where lawyers contest the degree of compatibility, those lawyers can and should interrogate the choice of $\alpha$ by selecting multiple $\alpha$ values and observing how sensitive the posterior distribution of the striking attorney's bias parameter is to the choice of $\alpha$ value. (Our protoype app permits this kind of testing of sensitivity in real time by providing multiple options for $\alpha$ values.) If the inference of bias is sensitive to the choice of $\alpha$ values, all else equal, then worries about incompatibility are warranted. As a result, the historical strike data, without more information, helps less to prove a _Batson_ violation than it would otherwise would.

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds that illegal bias is more likely than not (e.g., a _Batson_ violation), because no approach to estimating bias based on historical strike data has that ability.

In future work we hope to examine data-driven approaches to selecting $\alpha$; where the degree of incompatibility can be estimated and used to inform the choice of the degree of discounting to apply to historical data. Knowing how to quantify incompatibility and how to specify an appropriate prior for $\alpha$ is an area of active research [e.g., @Pawel2022]. In any case, any formulation for estimating $\alpha$ based on incompabitiliy between current and historical trials would require justification in court (or by practitioners).

@@

## CRediT Author Statement

@@

Here, we extend the discussion in section \@ref(prototype) of the paper by discussing the data collection and presenting additional illustrations of how a lawyer might use the prototype app.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make implicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection process. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences the bias detection (whether the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

In any case, evidence of such illegal bias may include data on the use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected such historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2022.

Without historical strike data, there is a higher risk of underpowered or inflated estimates of attorney strike bias, because of the low number of strikes per trial: typically from 6 to  15 strikes per party, depending on the jurisdiction [@nscs-jury2018]. Yet, complete pooling of current and historical strike data may be affected by the degree of incompatibility between current and historical strike data. Such incompatibility may occur if missing historical strike is missing not at random due to incomplete or inaccessible jury-selection records [@Grosso2017; @Wright2018].

Incompatibility may also occur if attorney strike bias depends on certain trial-level characteristics (e.g., defendant race, charge severity) that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then inferring strike bias in the current trial with a Black defendant may depend in part on whether we use all the historical data or only the subset of past trials with Black defendants.

As a result, for any given estimate of strike bias in a current trial, incompatibility between strike data from that current trial and strike data from past trials can lead to inflated type I error and reduced power [@Bennett2021]. Other methods to increase power may in turn require adjusting for incompatibility implicitly. For example, @Gastwirth2014 [p. 295-297] stratify strike data by trial and apply the Cochran-Mantel-Haenszel test, albeit after first selecting only historical strike data from past trials that are "similar" in their bias-salient characteristics to the current trial [p. 289, 297]. This initial step presumably aims to reduce incompatibility due to variation in those salient characteristics.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows for transparently adjusting for the assumed degree of incompatibility between current and historical trials. We proceed in three steps.  First, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Second, we use an attorney's strike data from past trials, generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Chen2000; @Ibrahim2015], which raises the likelihood of the historical strike data to a fixed power ($\alpha$), typically between zero and one, to down-weight the historical strike data in accord with expected incompatibility between the historical trials and the current trial. Zero denotes no weight to the historical data (for complete incompatibility) and one denotes equal weight (for complete compatibility). Third, we use this posterior distribution as an informative prior for the bias parameter in the current trial.

We demonstrate this approach with simulations in which we vary the degree of incompatibility between current and historical trial data, and test how well we can "detect" attorney strike bias, i.e., whether the credible interval for the bias parameter excludes zero, given different values of the power prior's alpha parameter. Then, we present a software prototype that encodes the same approach with actual historical strike data from a convenience sample of criminal cases from one court. With that data, the prototype shows how attorneys and others can use this approach to detect strike bias in real time during jury selection.

@@

```
To be sure, neither this nor any other statistical approach can automatically detect _Batson_ violations, because _Batson_ law permits any evidence of strike bias, not just strikes in past cases. And _Batson_ law  requires the trial judge to weigh all such evidence to decide whether strike bias likely exists (the _Batson_ violation). Still, _before_ accounting for other relevant evidence of strike bias, our approach can help attorneys detect strike bias, given current and historical strike data, the model of the relevant court's strike procedure, and a fixed value for the power prior's $\alpha$ parameter. After testing for sensitivity of bias detection to different fixed $\alpha$ values for the power prior [@Carvalho2021, p. 5252; @Ibrahim2015, p. 3734], attorneys can then combine the resulting inference with other evidence to decide whether to argue _Batson_; to prove the _Batson_ prima facie case; and to persuade the trial judge to find  a _Batson_ violation (Figure \@ref(fig:batsonworkflowfig)).

@@

Following others [e.g., @Kadane2018a; @Barrett2007; @Gastwirth2005, p. 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptory strikes. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

To evaluate the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. These simulations primarily show how much the size of the historical data affects our ability to detect bias and how much the power prior discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

@@

Bias detection will also be sensitive to the choice of the standard deviation of the initial prior ($\sigma$). The results above depend in part on our choice of a conservative initial prior for the bias parameter (Equation \@ref(eq:initialprior)). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, our model will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. As in the simulations, a normal distribution was assumed for the distribution of $b$. The length of the Markov chain was 110000 and the first 10000 iterations were dropped as burn-in. No thinning was performed, as the correlation was weak and convergence occured rapidly. Convergence was checked from both traceplots and the Gelman-Rubin statistic [@Gelman1992]. Computation times vary from 0.76 to 1.2 seconds, depending in the amount of data used (quicker with current data only _versus_ current plus historical data).

To show how lawyers might use this app in real cases in real time during jury selection, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. Using publicly-available court docket sheets for each case, we matched particular lawyers to these strikes and later de-identified them. For details on this dataset, see the Appendix.

@@

To illustrate, suppose the strike tally in Table 1 depicts the pattern of strikes in the present case with defense attorney Aaron Waelchi and prosecutor Lawrance Klocko V (both aliases for actual attorneys in the historical strike data). After the user enters this strike tally, the app initially displays two graphs -- one for the prosecution and the defense. Each graph depicts the prior density plot (colored light grey) and posterior density plot (blue and red for defense and prosecution, respectively) for the bias parameter (Figure \@ref(fig:figapp1)(a)). Here, the 80\% credible interval includes zero, indicating no credible basis to infer bias, given the current strike tally.

```{r figapp1, out.width='50%', fig.cap = "Screenshots of R-Shiny application showing density plots for race bias of prosecutor and defense attorney. Vertical dotted lines depict 80 pct. credible interval.", fig.align='center', fig.ncol = 2, fig.show = 'hold', fig.subcap=c("Current Strike Data only","Historical and current strike data")}

load_path1 <- here::here("figures","batson_app_screenshots", "batson_app_screenshot1.png")
load_path2 <- here::here("figures","batson_app_screenshots", "batson_app_screenshot2.png")

par(mfrow=c(2,1))
knitr::include_graphics(load_path1)
knitr::include_graphics(load_path2)

@@

In our illustration, we select the names of the prosecutor and defense attorney from their respective pull-down menus; and leave the weight setting to "Equal". The density plots update accordingly (Figure \@ref(fig:figapp1)(b)). Now the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case. In the Appendix, we present additional illustrations wherein we use the app to detect gender bias and where we assign less weight to the historical strike data.

```{r figapp2, fig.cap = "Density Plots with Historical Strike Data (Race). Screenshot of R-Shiny application show density plots based on historical and current strike data for prosecutor and defense attorney.", fig.align='center', eval = FALSE}

load_path2 <- here::here("figures","batson_app_screenshots", "batson_app_screenshot2.png")

knitr::include_graphics(load_path2)

@@

@@

Statistical methods for _Batson_ and similar challenges often use historical strike data to infer illegal strike bias. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, this approach makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not. Moreover, with a tool like our prototype app, attorneys can use the approach presented here in real time during jury selection to help them  decide whether to make a  _Batson_ challenge at all. If argued, the attorney can also introduce that posterior distribution as relevant evidence, along with any other admissible evidence, for presenting a prima facie case and for ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)).

## Selecting the Power Prior Value

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select for the power prior. Given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

Bias detection also increases with the number of historical trials and the number of strikes per trial. Higher values of $\alpha$ lead to higher bias detection rates across compatibility scenarios, especially in scenarios with a lower number of strikes per trial and few historical trials (top left panel of Figure \@ref(fig:figcmptbly)). The choice of $\alpha$ matters less for bias detection when bias is high; the number of strikes per trial is high; and the number of historical trials is high (lower right panel of Figure \@ref(fig:figcmptbly)). Overall, given high compatibility, increasing $\alpha$ will increase bias detection. But our simulations also show that, given high incompatibility, bias detection can be erroneously high when current bias ($b_{curr}$) is low; and bias detection can be erroneously low where current bias is high but bias in historical trials ($b_{hist}$) is low.

Where lawyers dispute the degree of incompatibility, they can and should interrogate the choice of $\alpha$ by selecting multiple $\alpha$ values and observing how sensitive the posterior distribution of the striking attorney's bias parameter is to the choice of $\alpha$ value. Our protoype app permits this kind of testing of sensitivity in real time by providing multiple options for $\alpha$ values. If the inference of bias is sensitive to the choice of $\alpha$ values, all else equal, then one should worry more about incompatibility and choose a lower $\alpha$ value accordingly.

This approach also applies when the unit of analysis is not the attorney, but the law firm or office. To illustrate, suppose prosecutors from the same office were all exposed to training that recommended striking Black jurors. If so, our approach allows for (but does not require) aggregating all those prosecutors' cases, i.e., acting as if all prosecutors from that office were a single individual prosecutor (e.g., the strike bias of the "district attorney's office"). Thereafter, any down-weighting adjustment (choosing $\alpha < 1$) depends on assumptions about how dissimilarities between the trial-level characteristics (e.g., defendant race, offense type) in the current trial and in past trials may still produce incompatibilty, notwithstanding the prosecutors' common exposure to the training. If none, one could assign equal weight to the historical data ($\alpha$ = 1). If one allows for some incompatibilty nonetheless, one could either subset the historical data (sacrificing power) to include only those past trials that share the same trial-characteristics as the bias-salient characteristics of current trial. Or one could keep all the historical data; choose an $\alpha$ value to match the degree of suspected incompatibility; and check for how sensitive any bias detection is to different $\alpha$ values.

In general, by their choice of $\alpha$, the lawyer explicitly and transparently weights the historical strike data in the analysis. This transparency thwarts lawyer efforts to strategically select $\alpha$ values. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides know that at any time the lawyers on the other side can always assess how sensitive any strike bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot alone determine whether a trial judge finds that illegal bias is more likely than not (i.e., a _Batson_ violation). Future work might examine how information-criteria or data-driven approaches for selecting $\alpha$ might apply in this context. Knowing how to quantify incompatibility and specify an appropriate prior for $\alpha$ is an area of active research [e.g., @Pawel2022].

## Modeling the Strike Procedure

The Bayesian approach here also requires a model to match the strike procedure of the court of interest. This choice of model can matter in at least three ways.  First, it can affect the missingness of historical strike data. For example, given the model we used (Equation \@ref(eq:model1)), we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike. Such information may be missing from jury selection records of past cases.

Second, the choice of model can affect prospective data collection not only by courts but also by courtroom observers. The public has a presumptive First Amendment right to attend jury voir dire in criminal cases. Similarly, a criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from voir dire [@press1984; @presley2010]. As a result, any person can often attend jury voir dire and collect relevant strike data based on what they observe in real time. Thus, such courtroom observers can prospectively collect strike data that they would have been less likely to obtain from court records after the fact.

Third, the choice of model matters if historical strike data for a particular attorney covers strikes exercised in multiple courts with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that can be applied to another court's procedure (e.g., the simultaneous strike example in section \@ref(simul-strikes) ). In other cases, however, one cannot apply the same model to courts with different jury selection practices. For example, the model we used here does not apply to strike data generated by a court's strike procedure in which prospective jurors appear before the attorneys one at a time to be either struck or seated on the jury.

@@

We thank the Associate Editor and two anonymous reviewers for their comments.

# CRediT Author Statement.

@@

Here, we extend the discussion in section \@ref(prototype) by discussing data collection and presenting additional illustrations of how a lawyer might use the prototype app.
